The Concurrent Calculi Formalisation Benchmark

### 2024-02-20 short-paper/short-paper.tex
minor rewrites to linearity

Linear typing systems allow for tracking the use of resources in a program. In
the case of typed process calculi (in particular, session types), linearity is
widely used for tracking if and how a channel is used to send or receive values.
This leads to substructural typing~\cite[Chapter 1]{Pierce2004-oq} with
resulting mechanisation difficulties --- \eg deciding how to \emph{split the
typing context} in a parallel composition.

The goal of our challenge on linear reasoning is proving a type safety theorem
for a process calculus with session types, by combining as usual subject
reduction with absence of errors.  We have chosen to only model linear (as
opposed to shared) channels. Inspired by Vasconcelos~\cite{Vasconcelos2012}, we
define a syntax where a restriction $(\nu ab)\ldots$ binds two dual names $a$
and $b$ as opposite endpoints of a same channel --- and their duality is
reflected in the type system. We model a simple notion of error: a well-typed
processes must never use dual channel endpoints in a non-dual way (\eg by
performing concurrent send/receive operations on a same endpoint, or two
concurrent send operations on both endpoints).
% The latter enforces two invariants, namely that no endpoint is
% used simultaneously by parallel processes and that the two endpoints of the same
% session have dual types.
The operational semantics is formulated as a standard reduction relation; hence,
proving subject reduction requires proving type preservation for structural
congruence.

@@

In mechanised meta-theory, addressing linearity means choosing an appropriate
representation of the linear typing context.  While a linear context is perhaps
best seen as a multi-set, most proof assistants have better support for list
processing.  The latter representation is intuitive, but may require
establishing a large number of technical lemmata that are orthogonal to the
problem under study (in our case, proving type safety for
session types).  In this space, several designs are possible: one can label
occurrences of resources to constrain their usage (\eg~\cite{CicconeP20}); some
authors instead impose a multiset structure over lists
(\eg~\cite{Danielsson12,ChaudhuriLR19}). Alternatively, contexts can be
implemented as finite maps (as in \cite{Castro2020}), whose operations are
sensitive to a linear discipline. In all these cases, the effort required to
develop the infrastructure is significant.
%
A different approach is to bypass the problem of context splitting by adopting familiar ideas from algorithmic linear type checking: this is known as \emph{typing with leftovers}, as shown in~\cite{DBLP:conf/forte/ZalakainD21}.
Whatever the choice, list-based encodings can be refined to be intrinsically-typed, if the proof assistant supports dependent types (see~\cite{Thiemann2019,CicconeP20,RouvoetPKV20}).
%
A completely different approach is to adopt a \emph{substructural} meta-logical framework: the framework itself handles resource distribution, and users need only map their linear operations to the ones offered by the framework.
The linear function space will then handle all context operations implicitly, including splitting and substitution.
One such framework is \emph{Celf}~\cite{Schack-Nielsen:IJCAR08} (see the encoding of session types in~\cite{Bock2016});
unfortunately, \emph{Celf} does not yet fully support the verification of meta-theoretic properties.
A compromise is the so called \emph{two-level} approach, \ie encoding a substructural specification logic in a mainstream proof assistant and then using that logic to state and prove linear properties (for a recent example, see~\cite{Felty:MSCS21}).

\subsection{Scope extrusion}
This challenge revolves around the mechanisation of explicit scope
extrusion, by which a process can send restricted names to another
process, as long as the restriction can safely be ``extruded'' (\ie,
expanded) to include the receiving process.  
The setting for this challenge is a ``classic'' untyped
\( \pi \)-calculus, where (unlike the calculi in the other challenges)
names can be sent and received, and bound by input constructs.  We define two
different semantics for our system:
\begin{enumerate}
\item A reduction system: this avoids explicit reasoning
about scope extrusion via structural congruence, allowing implementors to explore
alternative ways to encode the latter (\eg via context or via congruent refinements)
\item A (early) labelled transition system
\end{enumerate}
The objective of this
challenge is to prove that the two semantics are equivalent up to
structural congruence.

>>>>

Linear typing systems allow  tracking the use of resources in a program. In
the case of typed process calculi (in particular, session types), linearity is
widely used for checking if and how a channel is used to send or receive values.
This leads to substructural typing~\cite[Chapter 1]{Pierce2004-oq} with
resulting mechanisation difficulties --- \eg deciding how to \emph{split the
typing context} in a parallel composition.

The goal of our challenge on linear reasoning is proving a type safety theorem
for a process calculus with session types, by combining as usual subject
reduction with absence of errors.  We have chosen  only to model linear (as
opposed to shared) channels. Inspired by Vasconcelos~\cite{Vasconcelos2012}, we
define a syntax where a restriction $(\nu ab)\ldots$ binds two dual names $a$
and $b$ as opposite endpoints of a same channel --- and their duality is
reflected in the type system. We model a simple notion of error: a well-typed
processes must never use dual channel endpoints in a non-dual way (\eg by
performing concurrent send/receive operations on a same endpoint, or two
concurrent send operations on both endpoints).
% The latter enforces two invariants, namely that no endpoint is
% used simultaneously by parallel processes and that the two endpoints of the same
% session have dual types.
The operational semantics is formulated as a standard reduction relation; hence,
proving subject reduction requires proving type preservation for structural
congruence.

@@

In mechanised meta-theory, addressing linearity means choosing an appropriate
representation of the linear typing context.  While a linear context is perhaps
best seen as a multi-set, most proof assistants have better support for list
processing.  The latter representation is intuitive, but may require
establishing a large number of technical lemmata that are orthogonal to the
problem under study (in our case, proving type safety for
session types).  In this space, several designs are possible: one can label
occurrences of resources to constrain their usage (\eg~\cite{CicconeP20}); some
authors instead impose a multiset structure over lists
(\eg~\cite{Danielsson12,ChaudhuriLR19}). Alternatively, contexts can be
implemented as finite maps (as in \cite{Castro2020}), whose operations are
sensitive to a linear discipline. In all these cases, the effort required to
develop the infrastructure is significant.
%
One alternative strategy is to bypass the problem of context splitting
by adopting familiar ideas from algorithmic linear type checking. This
approach, referred to as ``typing with leftovers'', is exemplified
in~\cite{DBLP:conf/forte/ZalakainD21}.  Whatever the choice,
list-based encodings can be refined to be intrinsically-typed, if the
proof assistant supports dependent types
(see~\cite{Thiemann2019,CicconeP20,RouvoetPKV20}).
%
A radically different approach is to adopt a \emph{substructural}
meta-logical framework: the framework itself handles resource
distribution, and users need only map their linear operations to the
ones offered by the framework.  The linear function space will then
handle all context operations implicitly, including splitting and
substitution.  The only such framework is
\emph{Celf}~\cite{Schack-Nielsen:IJCAR08} (see the encoding of session
types in~\cite{Bock2016}); unfortunately, \emph{Celf} does not yet
fully support the verification of meta-theoretic properties.  A
compromise is the so called \emph{two-level} approach, \ie encoding a
substructural specification logic in a mainstream proof assistant and
then using that logic to state and prove linear properties (for a
recent example, see~\cite{Felty:MSCS21}).

\subsection{Scope extrusion}
This challenge revolves around the mechanisation of explicit scope
extrusion, by which a process can send restricted names to another
process, as long as the restriction can safely be extruded to include
the receiving process.  The setting for this challenge is a
``classic'' untyped \( \pi \)-calculus, where (unlike the calculi in
the other challenges) names can be sent and received, and bound by
input constructs.  We define two different semantics for our system:
\begin{enumerate}
\item A reduction system: this avoids explicit reasoning
about scope extrusion via structural congruence, allowing implementors to explore
alternative ways to encode the latter (\eg via context or via congruent refinements)
\item A (early) labelled transition system
\end{enumerate}
The objective of this
challenge is to prove that the two semantics are equivalent up to
structural congruence.
